\MakeUppercase{The Structure of $k$-Planar Graphs}

### 2019-07-09 k-planar.tex
David's work

\begin{abstract}
Dujmovi\'c~et~al.~(FOCS 2019) recently proved that every planar graph is a subgraph of the strong product of a graph of bounded treewidth and a path. This tool has been used to solve longstanding problems on queue layouts, non-repetitive colouring, and $p$-centred colouring. %, and graph encoding. 
We generalise this result for $k$-planar graphs, where a graph is \emph{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings. In particular, we prove that every $k$-planar graph is a subgraph of the strong product of a graph of treewidth $O(k^5)$ and a path. This is the first result of this type for a non-minor-closed class of graphs. It implies, amongst other results, that $k$-planar graphs have non-repetitive chromatic number upper-bounded by a function of $k$. All these results generalise for drawings of graphs on arbitrary surfaces. 
\end{abstract}
\end{titlepage}

@@

\tableofcontents
\newpage
\section{Introduction}

A graph is \emph{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings. Such graphs provide a natural generalisation of planar graphs, and are important in graph drawing research; see the recent bibliography on 1-planar graphs and the 140 references therein \citep{kobourov.liotta.ea:annotated}. The present paper studies the structure of $k$-planar graphs (and other more general classes). Various structural results about planar graphs generalise for $k$-planar graphs. For example, \citet{FP08} generalised the Lipton-Tarjan separator theorem to show that every $k$-planar graph with $n$ vertices has a balanced separator of order $O(\sqrt{kn})$.\note{PM}{I must be missing something.  The planarized graph $G'$ has $O(kn)$ vertices.  Let $S$ be the set of non-dummy vertices and take a separation $(A,B)$ of $G'$ with $|A\cap B|=O(\sqrt{kn})$, $|S\cap (A\setminus B)|\le 2|S|/3$ and $|S\cap (B\setminus A)|\le 2|S|/3$.  Now take $X=A\cap B$ plus the endpoints of every edge of $G$ that contains a dummy vertex in $A\cap B$.  $|X|\le 5|A\cap B|=O(\sqrt{kn})$ and $X$ is a separator of $G$.}

@@

\begin{thm}
\label{kPlanarBasic}
Every $k$-planar graph is a subgraph of $H\boxtimes P$, for some graph $H$ of treewidth $O(k^5)$ and for some path $P$. 
\end{thm}

@@

We prove the following theorem.

@@

\cref{ShortcutProduct,AddDummy} immediately imply \cref{kPlanarBasic} with $O(k^{11})$ instead of $O(k^5)$. Some further observations reduce the bound to $O(k^5)$; see \cref{Tools}. All of these results  generalise for graphs that have a drawing on an arbitrary surface with a bounded number of crossings per edge.

@@

\begin{thm}
\label{kPowerBasic}
For every planar graph $G$ with maximum degree $\Delta$ and for every integer $k\geq 1$, $G^k$ is a subgraph of $H\boxtimes P$, for some graph $H$ of treewidth at most $O( k^7 \Delta^{k})$ and for some path $P$. \note{DW}{Improve to $k^6\Delta^{k}$?}
\end{thm}
\note{PM}{TODO: Work out the treewidth more carefully. The use of big-O is here difficult to decipher (and probably incorrect).}

@@

\cref{Examples} presents several examples of $k$-planar graphs, including graph powers, map graphs, string graphs, and nearest neighbour graphs. All of the above results are applicable to these examples.

@@

In this paper, all graphs are finite and undirected. Unless specifically mentioned otherwise, all graphs are also simple. For any graph $G$ and any set $S$ (typically $S\subseteq V(G)$) we use $G[S]$ to denote the graph with vertex set $V(G)\cap S$ and edge set $\{uv\in E(G) : u,v\in S\}$.  We use $G-S$ as a shorthand for $G[V(G)\setminus S]$. We use $G'\subseteq G$ to denote subgraph containment; that is, $V(G')\subseteq V(G)$ and $E(G')\subseteq E(G)$.

@@

We now define two concepts used in the statements of \cref{PlanarBasic,kPlanarBasic}: strong products and treewidth. The \emph{strong product} of graphs $A$ and $B$, denoted by $A\boxtimes B$, is the graph with vertex set $V(A)\times V(B)$, where distinct vertices $(v,x),(w,y)\in V(A)\times V(B)$ are adjacent if: 
\begin{compactitem}
\item  $v=w$ and $xy\in E(B)$, or 
\item  $x=y$ and $vw\in E(A)$, or  
\item  $vw\in E(A)$ and $xy\in E(B)$. 
\end{compactitem}
A \emph{tree-decomposition} $\mathcal{T}$ of a graph $G$ consists of a tree $T$ and a collection $\mathcal{T}=(B_x:x\in V(T))$ of subsets of $V(G)$ indexed by the nodes of $K$ such that:
\begin{compactenum}[(i)]
\item for every $vw\in E(G)$, there exists some node $x\in V(T)$ with $v,w\in B_x$; and 
\item for every $v\in V(G)$, the induced subgraph $T[v] := T[\{x: v\in B_x\}]$ is connected.  
\end{compactenum}
The \emph{width} of the tree-decomposition $\mathcal{T}$ is $\max\{|B_x|:x\in V(T)\}-1$.  The \emph{treewidth} $\tw(G)$ of a graph $G$ is the minimum width of a tree-decomposition of $G$.  Treewidth is the standard measure of how similar a graph is to a tree. Indeed, a connected graph has treewidth 1 if and only if it is a tree. Treewidth is of fundamental importance in structural and algorithmic graph theory; see \citep{Reed03,HW17,Bodlaender-TCS98} for surveys.

While strong products enable an easy statement of \cref{PlanarBasic,kPlanarBasic}, to prove such theorems it is helpful to work with layerings and partitions, which we now introduce. A \emph{layering} of a graph $G$ is a sequence $\mathcal{L}=\langle V_0,V_1,\ldots\rangle$ such that $\{V_0,V_1,\ldots\}$ is a partition of $V(G)$ and for every edge $vw\in E(G)$, if $v\in V_i$ and $w\in V_j$ then $|j-i|\leq 1$.  For any partition $\mathcal{P}=\{S_1,\ldots,S_p\}$ of $V(G)$, a \emph{quotient graph} $H=G/\mathcal{P}$ has a $p$-element vertex set $V(H)=\{x_1,\ldots,x_p\}$ and $x_ix_j\in E(H)$ if and only if there exists an edge $vw\in E(G)$ such that $v\in S_i$ and $w\in S_j$. To highlight the importance of the quotient graph $H$, we call $\mathcal{P}$ an \emph{$H$-partition} and write this concisely as $\mathcal{P}=\{S_x : x\in V(H)\}$ so that each element of $\mathcal{P}$ is indexed by the vertex it creates in $H$.\footnote{An alternative definition of a layering of $G$ is an $H$-partition of $G$ where $H$ is a path.}  A layered $H$-partition $(\mathcal{L},\mathcal{P})$ of a graph $G$ consists of a layering $\mathcal{L}$ of $G$ and an $H$-partition $\mathcal{P}$ of $G$. The \emph{layered width} of $(\mathcal{L},\mathcal{P})$ is $\max\{|L\cap P|: L\in\mathcal{L},\, P\in\mathcal{P}\}$.

@@

\begin{thm}[\citep{dujmovic.joret.ea:planar}]
\label{PlanarPartition}
Every planar graph has:
\begin{compactitem}
\item an $H$-partition of layered width $1$ for some planar graph $H$ of treewidth at most $8$, and
\item an $H$-partition of layered width $3$ for some planar graph $H$ of treewidth at most $3$.
\end{compactitem}
\end{thm}
Their proof is constructive and gives a simple quadratic-time algorithm for finding these partitions.

@@

\citet{dujmovic.joret.ea:planar} generalised \cref{SurfacePartition} further to show that for every apex\footnote{A graph $M$ is a \textit{minor} of a graph $G$ if a graph isomorphic to $M$ can be obtained from a subgraph of $G$ by contracting edges. A class $\mathcal{G}$ of graphs is \emph{minor-closed} if for every graph $G\in\mathcal{G}$, every minor of $G$ is in $\mathcal{G}$. A minor-closed class is \emph{proper} if it is not the class of all graphs. For example, for fixed $g\geq 0$, the class of graphs with Euler genus at most $g$ is a proper minor-closed class. A graph $G$ is $t$-apex if it contains a set $A$ of at most $t$ vertices such that $G-A$ is planar. A 1-apex graph is \emph{apex}.  A minor-closed class $\mathcal{G}$ is apex-minor-free if some apex graph is not in $\mathcal{G}$.}
 graph $X$, there is a constant $c$ such that every $X$-minor-free graph is a subgraph of $H\boxtimes P$ for some graph $H$ of treewidth at most $c$. Apex-minor-free graphs are the largest minor-closed class that admit bounded layered partitions~\citep{dujmovic.joret.ea:planar}. However, the family of $k$-planar graphs is not minor-closed.  A graph $G'$ obtained from a $k$-planar graph $G$ by edge deletions and edge contractions may or may not be $k$-planar. Indeed, \citet{dujmovic.eppstein.ea:structure} construct 1-planar graphs that contain arbitrarily large complete graph minors. Our results for $k$-planar graphs are the first of this type for a non-minor-closed class.

@@

\cref{PlanarPartition,modern-major-general} imply \cref{k-planar} with the weaker bound of $(2\times 3)((k+1)^3+3(k+1))=6k^3 + 18k^2 + 36k+24$ on the layered width. We prove the stronger bound in \cref{sec-k-planar}. 
%Also note that \cref{k-planar,MakeWidth1} imply \cref{kPlanarBasic} with the $O(k^5)$ bound on the treewidth of $H$, as promised. 
% PM: I removed the last line because it's already stated in the line preceding the theorem

@@

\begin{obs}
\label{gAddDummy}
Every $(g,k)$-planar graph is isomorphic to $G^\PP$ for some graph $G$ of Euler genus $g$ and some $(k+1,2)$-shortcut system $\PP$ for $G$. 
\end{obs}

@@

Finally, we state the following corollary of \cref{modern-major-general,MakeWidth1,PartitionProduct} and the above-mentioned theorem for apex-minor-free graphs.  This is the strongest result that follows from our work and that of \citet{dujmovic.joret.ea:planar}.

@@

\note{DW}{should this subsection be moved to the end of \cref{Tools}?}\note{PM}{No, I don't think so. The information given in Section 2 isn't detailed enough to compare with layered treewidth versus layered partitions.}

@@

For a rooted tree $T$ rooted at some node $x_0\in V(T)$, we we say that a node $a\in V(T)$ is a \emph{$T$-ancestor} of $x\in V(T)$ (and $x$ is a \emph{$T$-descendant} of $a$) if $a$ is a vertex of the path, in $T$, from $x_0$ to $x$.  Note that each node $x\in v(T)$ is a $T$-ancestor and $T$-descendant of itself.  We say that a $T$-ancestor $a\in V(T)$ of $x\in V(T)$ is a \emph{strict} $T$-ancestor of $x$ if $a\neq x$. For a rooted tree $T$ and any node $x\in V(T)$, let  
\[T_x := T[\{y\in V(T):\mbox{$x$ is a $T$-ancestor of $y$}\}] \]
be the maximal subtree of $T$ rooted at $x$.

@@

\begin{lem}\lemlabel{nice-decomposition}
  For any graph $H$ of treewidth $t$, there is a rooted tree $T$ with $V(T)=V(H)$ and a $T$-decomposition $(B_x:x\in V(T))$ of $H$ that has following additional properties:  
  \begin{compactenum}[(T1)]
    % \item\tlabel{rooted}\tlabel{first} $T$ is rooted at some node $x_0\in V(T)$ with $B_{x_0}=\emptyset$.
    % \item\tlabel{node-set} $V(T)= V(H)$.
    % \item\tlabel{diff} For each edge $xy\in E(T)$, $|B_x\ominus B_y|\le 1$.
    \item\tlabel{subtree-root} For each node $x\in V(H)$, the subtree $T[x]:=T[\{y\in V(T):x\in B_y\}]$ is rooted at $x$.
    \item\tlabel{ancestor-edge}\tlabel{last} For each edge $xy\in E(H)$, one of $x$ or $y$ is a $T$-ancestor of the other.
  \end{compactenum}
\end{lem}

@@

For the remainder of this section, $G$ is a graph and $(\mathcal{L}:=\langle V_1,\ldots,V_h\rangle,\mathcal{Y}:=(Y_x: x\in V(H)))$ is a layered partition of $G$ of layered width $\ell$ where the treewidth of the quotient graph $H=G/\mathcal{Y}$ is at most $t$.  In addition, $\mathcal{T}:=(B_x:x\in V(T))$ is a tree-decomposition of $H$ obtained by applying \lemref{nice-decomposition} to $H$.  The following lemma shows how to interpret the pair $(\mathcal{Y},\mathcal{T})$ as a hierarchical decomposition of $G$.  Refer to \cref{fig:generalized-tripod}

@@

@@

\begin{proof}
  For the tree-decomposition $(C_x:x\in V(T))$ of $J$ we use the same tree $T$ used in the tree-decomposition $(B_x:x\in V(T))$ of $H$. For each node $x$ of $T$, $C_x$ contains $x$ as well as every $T$-ancestor $a$ of $x$ such that $J$ contains an edge $ax'$ where $x$ is a $T$-ancestor of $x'$ (including the possibility that $x=x'$).
  \clmref{i-ancestor} ensures that, for every edge $ax'\in E(J)$, $a,x'\in C_{x'}$ and the connectivity of $T[a]:=T[\{x\in V(T):a\in C_x]$ is obvious.  Therefore $(C_x:x\in V(T))$ is indeed a tree-decomposition of $J$.  What remains is to bound the size, $\max\{|C_x|:x\in V(T)\}$, of the largest bag.

@@

\begin{proof}[Proof of \cref{modern-major-general}] 
  The proof is almost immediate, except that the layering $\mathcal{L}$ of $G$ may not be a valid layering of $G^{\mathcal{P}}$.  In particular, $G^{\mathcal{P}}$ may contain an edge $vw$ with $v\in V_i$ and $w\in V_{i+j}$ for any $j\in\{0,\ldots,k\}$.  To resolve this, we use a new layering $\mathcal{L}':=\langle V_0',\ldots,V_h'\rangle$ in which $V_i'=\bigcup_{j=ki}^{ki+k-1} V_i$.  This increases the layered width given by \clmref{general-width} from $d\ell(k^2+3)$ to $d\ell(k^3+3k)$.
\end{proof}

@@

In this section we show that the constants in \cref{modern-major-general} can be improved for the important special case of $1$-planar graphs. Let $G$ be a $k$-plane graph.  We will assume, for ease of exposition, that any point $p\in\R^2$ is involved in at most one crossing $(p,vw,xy)$ of $G$. This assumption is not critical since it can be enforced by a slight deformation of the edges of $G$ and the resulting (deformed) graph is also $k$-plane.

Let $G_0$ be the plane graph obtained by adding a dummy vertex at each crossing in $G$. In this way, each edge $vw\in E(G)$ corresponds naturally to a path $P_{vw}$ of length at most $k+1$ in $G_0$.  Let $\PP := \{P_{vw}: vw\in E(G)\}$. Observe that $\PP$ is a $(k+1,2)$-shortcut system for $G_0$ and that $G_0^{\mathcal{P}}\supseteq G$.  Specifically, $G_0^{\mathcal{P}}$ contains every edge and vertex of $G$ as well as the dummy vertices in $V(G_0)\setminus V(G)$ and their incident edges.

Since $G_0$ is planar, the second part of \cref{PlanarPartition} implies that $G_0$ has an $H$-partition of layered width 3 for some planar graph $H$ of treewidth at most 3.  Applying \cref{modern-major-general} to $G_0$ and $\mathcal{P}$ immediately implies that $G$ (an arbitrary $k$-planar graph) has a layered $H$-partition of layered width $6((k+1)^3+3(k+1))$ for some graph $H$ of treewidth at most $\binom{k+4}{3}-1$.

>>>>

\begin{abstract}
Dujmovi\'c~et~al.~(FOCS 2019) recently proved that every planar graph is a subgraph of the strong product of a graph of bounded treewidth and a path. This tool has been used to solve longstanding problems on queue layouts, non-repetitive colouring, and $p$-centred colouring. %, and graph encoding. 
We generalise this result for $k$-planar graphs, where a graph is \emph{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings. In particular, we prove that every $k$-planar graph is a subgraph of the strong product of a graph of treewidth $O(k^5)$ and a path. This is the first result of this type for a non-minor-closed class of graphs. It implies, amongst other results, that $k$-planar graphs have non-repetitive chromatic number upper-bounded by a function of $k$. All these results generalise for drawings of graphs on arbitrary surfaces. In fact, we work in a much more general setting based on so-called shortcut systems that are of independent interest. 
\end{abstract}
\end{titlepage}

@@

\tableofcontents
\newpage
\section{Introduction}
\label{Introduction}

A graph is \emph{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings. Such graphs provide a natural generalisation of planar graphs, and are important in graph drawing research; see the recent bibliography on 1-planar graphs and the 140 references therein \citep{kobourov.liotta.ea:annotated}. The present paper studies the structure of $k$-planar graphs (and other more general classes). Various structural results about planar graphs generalise for $k$-planar graphs. For example, \citet{FP08} generalised the Lipton-Tarjan separator theorem to show that every $k$-planar graph with $n$ vertices has a balanced separator of order $O(\sqrt{kn})$.\note{PM}{I must be missing something.  The planarized graph $G'$ has $O(kn)$ vertices.  Let $S$ be the set of non-dummy vertices and take a separation $(A,B)$ of $G'$ with $|A\cap B|=O(\sqrt{kn})$, $|S\cap (A\setminus B)|\le 2|S|/3$ and $|S\cap (B\setminus A)|\le 2|S|/3$.  Now take $X=A\cap B$ plus the endpoints of every edge of $G$ that contains a dummy vertex in $A\cap B$.  $|X|\le 5|A\cap B|=O(\sqrt{kn})$ and $X$ is a separator of $G$.} \note{DW}{I think you have just found a new proof of this fact. Note that you are actually using that $G'$ has treewidth $O(\sqrt{kn})$ and then applying the Robertson-Seymour separator lemma for graphs of given treewidth (so that $|S\cap (A\setminus B)|\le 2|S|/3$ and $|S\cap (B\setminus A)|\le 2|S|/3$). }

@@

\begin{thm}
\label{kPlanarBasic}
Every $k$-planar graph is a subgraph of $H\boxtimes P$, for some graph $H$ of treewidth $(18k^3+30k)\binom{k+4}{3}-1$ and for some path $P$. 
\end{thm}

@@

We prove the following theorem that shows if a graph $G$ is a subgraph of $H\boxtimes P$ and $\PP$ is a shortcut system for $G$, then $G^{\mathcal{P}}$ is a subgraph of $J\boxtimes P$, where the treewidth of $J$ is bounded by a function of the treewidth of $H$.

@@

\cref{ShortcutProduct,AddDummy} immediately imply \cref{kPlanarBasic} with $O(k^{11})$ instead of $O(k^5)$. Some further observations reduce the bound to $O(k^5)$; see \cref{Tools}.

@@

\begin{thm}
\label{kPowerBasic}
For every planar graph $G$ with maximum degree $\Delta$ and for every integer $k\geq 1$, $G^k$ is a subgraph of $H\boxtimes P$, for some graph $H$ of treewidth at most 
$\Delta^{k} (2k^4+6k^2)\binom{k+8}{8}-1$. 
\end{thm}

Note that the dependence on $k$ in the bound on the treewidth of $H$ in \cref{kPowerBasic} is improved from $O(k^{12})$ to $O(k^7)$ in \cref{Powers}.

@@

\cref{Examples} presents several examples of graph classes that can be obtained from a shortcut system applied to a planar graphs, including graph powers, map graphs, string graphs, and nearest neighbour graphs. All of the above results are applicable to these examples.

All of the above results apply, where instead of planar graphs, we consider graphs of bounded Euler genus.

@@

In this paper, all graphs are finite and undirected. Unless specifically mentioned otherwise, all graphs are also simple. For any graph $G$ and any set $S$ (typically $S\subseteq V(G)$), let $G[S]$  denote the graph with vertex set $V(G)\cap S$ and edge set $\{uv\in E(G) : u,v\in S\}$.  We use $G-S$ as a shorthand for $G[V(G)\setminus S]$. We use $G'\subseteq G$ to denote subgraph containment; that is, $V(G')\subseteq V(G)$ and $E(G')\subseteq E(G)$.

@@

We now define two concepts used in the theorems in \cref{Introduction}: strong products and treewidth. The \emph{strong product} of graphs $A$ and $B$, denoted by $A\boxtimes B$, is the graph with vertex set $V(A)\times V(B)$, where distinct vertices $(v,x),(w,y)\in V(A)\times V(B)$ are adjacent if: 
\begin{compactitem}
\item  $v=w$ and $xy\in E(B)$, or 
\item  $x=y$ and $vw\in E(A)$, or  
\item  $vw\in E(A)$ and $xy\in E(B)$. 
\end{compactitem}
A \emph{tree-decomposition} $\mathcal{T}$ of a graph $G$ consists of a tree $T$ and a collection $\mathcal{T}=(B_x:x\in V(T))$ of subsets of $V(G)$ indexed by the nodes of $K$ such that:
\begin{compactenum}[(i)]
\item for every $vw\in E(G)$, there exists some node $x\in V(T)$ with $v,w\in B_x$; and 
\item for every $v\in V(G)$, the induced subgraph $T[v] := T[\{x: v\in B_x\}]$ is connected.  
\end{compactenum}
The \emph{width} of the tree-decomposition $\mathcal{T}$ is $\max\{|B_x|:x\in V(T)\}-1$.  The \emph{treewidth} $\tw(G)$ of a graph $G$ is the minimum width of a tree-decomposition of $G$.  Treewidth is the standard measure of how similar a graph is to a tree. Indeed, a connected graph has treewidth 1 if and only if it is a tree. Treewidth is of fundamental importance in structural and algorithmic graph theory; see \citep{Reed03,HW17,Bodlaender-TCS98} for surveys.

While strong products enable an easy statement of the theorems in \cref{Introduction}, to prove such results it is helpful to work with layerings and partitions, which we now introduce. A \emph{layering} of a graph $G$ is a sequence $\mathcal{L}=\langle V_0,V_1,\ldots\rangle$ such that $\{V_0,V_1,\ldots\}$ is a partition of $V(G)$ and for every edge $vw\in E(G)$, if $v\in V_i$ and $w\in V_j$ then $|j-i|\leq 1$.  For any partition $\mathcal{P}=\{S_1,\ldots,S_p\}$ of $V(G)$, a \emph{quotient graph} $H=G/\mathcal{P}$ has a $p$-element vertex set $V(H)=\{x_1,\ldots,x_p\}$ and $x_ix_j\in E(H)$ if and only if there exists an edge $vw\in E(G)$ such that $v\in S_i$ and $w\in S_j$. To highlight the importance of the quotient graph $H$, we call $\mathcal{P}$ an \emph{$H$-partition} and write this concisely as $\mathcal{P}=\{S_x : x\in V(H)\}$ so that each element of $\mathcal{P}$ is indexed by the vertex it creates in $H$.\footnote{An alternative definition of a layering of $G$ is an $H$-partition of $G$ where $H$ is a path.}  A layered $H$-partition $(\mathcal{L},\mathcal{P})$ of a graph $G$ consists of a layering $\mathcal{L}$ of $G$ and an $H$-partition $\mathcal{P}$ of $G$. The \emph{layered width} of $(\mathcal{L},\mathcal{P})$ is $\max\{|L\cap P|: L\in\mathcal{L},\, P\in\mathcal{P}\}$.

\note{DW}{I  prefer to not define ``layered $H$-partition''. Just define the ``layered width'' of an $H$-partition, which is what we did in \citep{dujmovic.joret.ea:planar}. There are many places where we write ``$H$-partition with layered width'' without writing ``layered $H$-partition''. I prefer it this way.}

@@

\begin{thm}[\citep{dujmovic.joret.ea:planar}]
\label{PlanarPartition}
Every planar graph has:
\begin{compactenum}[(a)]
\item an $H$-partition of layered width $1$ for some planar graph $H$ of treewidth at most $8$, and
\item an $H$-partition of layered width $3$ for some planar graph $H$ of treewidth at most $3$.
\end{compactenum}
\end{thm}
Their proof is constructive and gives a simple quadratic-time algorithm for finding these partitions.

@@

\citet{dujmovic.joret.ea:planar} generalised \cref{SurfacePartition} further as follows.\footnote{A graph $M$ is a \textit{minor} of a graph $G$ if a graph isomorphic to $M$ can be obtained from a subgraph of $G$ by contracting edges. A class $\mathcal{G}$ of graphs is \emph{minor-closed} if for every graph $G\in\mathcal{G}$, every minor of $G$ is in $\mathcal{G}$. A minor-closed class is \emph{proper} if it is not the class of all graphs. For example, for fixed $g\geq 0$, the class of graphs with Euler genus at most $g$ is a proper minor-closed class. A graph $G$ is $t$-apex if it contains a set $A$ of at most $t$ vertices such that $G-A$ is planar. A 1-apex graph is \emph{apex}.  A minor-closed class $\mathcal{G}$ is apex-minor-free if some apex graph is not in $\mathcal{G}$.}

\begin{thm}[\citep{dujmovic.joret.ea:planar}] 
\label{ApexMinorFree}
For every apex graph $X$, there exists $c\in\mathbb{N}$ such that every $X$-minor-free graph $G$ has
an $H$-partition with layered width 1 such that $H$ has treewidth at most $c$. Equivalently, $G\subseteq H\boxtimes P$ for some path $P$. 
\end{thm}

Apex-minor-free graphs are the largest minor-closed class that admit bounded layered partitions~\citep{dujmovic.joret.ea:planar}. However, the family of $k$-planar graphs is not minor-closed.  A graph $G'$ obtained from a $k$-planar graph $G$ by edge deletions and edge contractions may or may not be $k$-planar. Indeed, \citet{dujmovic.eppstein.ea:structure} construct 1-planar graphs that contain arbitrarily large complete graph minors. Our results for $k$-planar graphs are the first of this type for a non-minor-closed class.

@@

\cref{PlanarPartition,modern-major-general} imply \cref{k-planar} with the weaker bound of $(2\times 3)((k+1)^3+3(k+1))=6k^3 + 18k^2 + 36k+24$ on the layered width. 
\note{DW}{I would just write, ``\cref{PlanarPartition,modern-major-general} imply \cref{k-planar} with a weaker bound on the layered width.}
We prove the stronger bound in \cref{sec-k-planar}. 
%Also note that \cref{k-planar,MakeWidth1} imply \cref{kPlanarBasic} with the $O(k^5)$ bound on the treewidth of $H$, as promised. 
% PM: I removed the last line because it's already stated in the line preceding the theorem

@@

\begin{obs}
\label{gAddDummy}
Every $(g,k)$-planar graph is isomorphic to $G^\PP$ for some graph $G$ of Euler genus at most $g$ and some $(k+1,2)$-shortcut system $\PP$ for $G$. 
\end{obs}

@@

Finally, we state the following corollary of \cref{ApexMinorFree,modern-major-general,MakeWidth1,PartitionProduct}. This is the strongest result that follows from our work and that of \citet{dujmovic.joret.ea:planar}.

@@

%\note{DW}{should this subsection be moved to the end of \cref{Tools}?}\note{PM}{No, I don't think so. The information given in Section 2 isn't detailed enough to compare with layered treewidth versus layered partitions.}

@@

For a tree $T$ rooted at some node $x_0\in V(T)$, we we say that a node $a\in V(T)$ is a \emph{$T$-ancestor} of $x\in V(T)$ (and $x$ is a \emph{$T$-descendant} of $a$) if $a$ is a vertex of the path, in $T$, from $x_0$ to $x$.  Note that each node $x\in v(T)$ is a $T$-ancestor and $T$-descendant of itself.  We say that a $T$-ancestor $a\in V(T)$ of $x\in V(T)$ is a \emph{strict} $T$-ancestor of $x$ if $a\neq x$. For a rooted tree $T$ and any node $x\in V(T)$, let  
\[T_x := T[\{y\in V(T):\mbox{$x$ is a $T$-ancestor of $y$}\}] \]
be the maximal subtree of $T$ rooted at $x$.

@@

\begin{lem}\lemlabel{nice-decomposition}
  For every graph $H$ of treewidth $t$, there is a rooted tree $T$ with $V(T)=V(H)$ and a $T$-decomposition $(B_x:x\in V(T))$ of $H$ that has following additional properties:  
  \begin{compactenum}[(T1)]
    % \item\tlabel{rooted}\tlabel{first} $T$ is rooted at some node $x_0\in V(T)$ with $B_{x_0}=\emptyset$.
    % \item\tlabel{node-set} $V(T)= V(H)$.
    % \item\tlabel{diff} For each edge $xy\in E(T)$, $|B_x\ominus B_y|\le 1$.
    \item\tlabel{subtree-root} For each node $x\in V(H)$, the subtree $T[x]:=T[\{y\in V(T):x\in B_y\}]$ is rooted at $x$.
    \item\tlabel{ancestor-edge}\tlabel{last} For each edge $xy\in E(H)$, one of $x$ or $y$ is a $T$-ancestor of the other.
  \end{compactenum}
\end{lem}

@@

For the remainder of this section, $G$ is a graph and $(\mathcal{L}:=\langle V_1,\ldots,V_h\rangle,\mathcal{Y}:=(Y_x: x\in V(H)))$ is a layered partition of $G$ of layered width $\ell$ where the treewidth of the quotient graph $H=G/\mathcal{Y}$ is at most $t$.  In addition, $\mathcal{T}:=(B_x:x\in V(T))$ is a tree-decomposition of $H$ obtained by applying \lemref{nice-decomposition} to $H$.  The following lemma shows how to interpret the pair $(\mathcal{Y},\mathcal{T})$ as a hierarchical decomposition of $G$; refer to \cref{fig:generalized-tripod}.

@@

\note{W}{I think \clmref{generalized-tripod} should be presented as a stand-alone lemma. Then we repeat the statement of \cref{modern-major-general} followed by $\\begin \{proof\} .... \\end\{proof\}$.}

\note{DW}{Is some sort of converse to \clmref{generalized-tripod}  also true? \clmref{generalized-tripod}  shows that if a graph $G$ has a layered partition with all the nice properties, then $G$ has this generalised tripod partition. Isn't also true that if $G$ has a generalised tripod partition, then $G$ has a layered partition with all the nice properties. The proof is basically in \citep{dujmovic.joret.ea:planar}, right?}

@@

\begin{proof}
  For the tree-decomposition $(C_x:x\in V(T))$ of $J$ we use the same tree $T$ used in the tree-decomposition $(B_x:x\in V(T))$ of $H$. For each node $x$ of $T$, $C_x$ contains $x$ as well as every $T$-ancestor $a$ of $x$ such that $J$ contains an edge $ax'$ where $x$ is a $T$-ancestor of $x'$ (including the possibility that $x=x'$).
  \clmref{i-ancestor} ensures that, for every edge $ax'\in E(J)$, $a,x'\in C_{x'}$ and the connectivity of $T[a]:=T[\{x\in V(T):a\in C_x]$ is obvious.  Therefore $(C_x:x\in V(T))$ is indeed a tree-decomposition of $J$.  It remains to bound the size of each bag $C_x$.

@@

\begin{proof}[Proof of \cref{modern-major-general}] 
  The proof is almost immediate from \clmref{general-width} and \clmref{general-bag-size}, except that the layering $\mathcal{L}$ of $G$ may not be a valid layering of $G^{\mathcal{P}}$.  In particular, $G^{\mathcal{P}}$ may contain an edge $vw$ with $v\in V_i$ and $w\in V_{i+j}$ for any $j\in\{0,\ldots,k\}$.  To resolve this, we use a new layering $\mathcal{L}':=\langle V_0',\ldots,V_h'\rangle$ in which $V_i'=\bigcup_{j=ki}^{ki+k-1} V_i$.  This increases the layered width given by \clmref{general-width} from $d\ell(k^2+3)$ to $d\ell(k^3+3k)$.
\end{proof}

@@

This section shows that the constants in \cref{modern-major-general} can be improved in the case of $k$-planar graphs. Let $G$ be a $k$-plane graph.  We will assume, for ease of exposition, that any point $p\in\R^2$ is involved in at most one crossing $(p,vw,xy)$ of $G$. This assumption is not critical since it can be enforced by a slight deformation of the edges of $G$ and the resulting (deformed) graph is also $k$-plane.

As in the proof of \cref{AddDummy}, let $G_0$ be the plane graph obtained by adding a dummy vertex at each crossing in $G$. In this way, each edge $vw\in E(G)$ corresponds naturally to a path $P_{vw}$ of length at most $k+1$ in $G_0$.  Let $\PP := \{P_{vw}: vw\in E(G)\}$. Observe that $\PP$ is a $(k+1,2)$-shortcut system for $G_0$ and that $G_0^{\mathcal{P}}\supseteq G$.  Specifically, $G_0^{\mathcal{P}}$ contains every edge and vertex of $G$ as well as the dummy vertices in $V(G_0)\setminus V(G)$ and their incident edges.

Since $G_0$ is planar,  \cref{PlanarPartition}(b) implies that $G_0$ has an $H$-partition of layered width 3 for some planar graph $H$ of treewidth at most 3.  Applying \cref{modern-major-general} to $G_0$ and $\mathcal{P}$ immediately implies that $G$ (an arbitrary $k$-planar graph) has a layered $H$-partition of layered width $6((k+1)^3+3(k+1))$ for some graph $H$ of treewidth at most $\binom{k+4}{3}-1$.

We can reduce the layered width of the $H$-partition of $G$ from $O(k^3)$ to $O(k^2)$ by observing that the dummy vertices in $V(G_0)\setminus V(G)$ do not contribute to the layered width of this partition.  In this setting, the proof of \cref{clm:general-width} is simpler since each vertex $w\in Y_x$ contributes at most two vertices to $V_i\cap Y_x$.  More precisely, each path $P\in\mathcal{P}$ containing an internal (dummy) vertex $w\in Y_x\cap (V_{i-j}\cup V_{i+j})$ contributes:
\begin{compactitem}
  \item  at most two vertices to $S_x\cap V_i$ for $j\in\{0,\ldots,\floor{(k+1)/2}\}$;
  \item at most one vertex to $S_x\cap V_j$ for $j\in\{\floor{(k+1)/2},\ldots,k+1\}$; 
  \item no vertices to $S_x\cap V_j$ for $j > k+1$.
\end{compactitem}
\note{DW}{do we need the bulletted list here?}
Redoing the calculation at the end of the proof of \cref{clm:general-width} then yields 
\begin{align*}
    |S_x\cap Y_i| &\le d\ell\left( 
       2 
       + 4\left\lfloor\frac{k+1}{2}\right\rfloor 
       + 2\left(k+1-\left\lfloor\frac{k+1}{2}\right\rfloor\right)
       \right)\\
       & =
       d\ell\left(
         2k + 2\left\lfloor\frac{k+1}{2}\right\rfloor + 4
       \right) \\
       & \le
       d\ell(3k+5) = 18k+30 \enspace .
\end{align*}    
With this change, the layered width of the partition given by \cref{modern-major-general} becomes $18k^2+30k$.  This establishes \cref{k-planar}.

Exactly the same approach using \cref{SurfacePartition}(b) instead of \cref{PlanarPartition}(b) establishes 
\cref{g-k-planar} for $(g,k)$-planar graphs.
